---
title: "Is U.S. Severe Weather Becoming More Volatile?"
author: "Benjamin Hobbs"
format: dashboard
server: shiny
---

```{r}
#| context: setup
#| message: false

library(tidyverse)
library(lubridate)
library(rvest)
library(shiny)

# Path for the combined cache file
combined_cache = file.path("data", "raw", "storm_events_raw_combined.csv")

# Load from the cache if it already exists, otherwise try to download
if (file.exists(combined_cache)) {
  raw_combined = read_csv(combined_cache,
                          col_types = cols(.default = col_character()),
                          show_col_types = FALSE)
} else {
  # This is the webpage that contains the files that I need, I need to
  # webscrape to get these files in this session
  base_url = "https://www.ncei.noaa.gov/pub/data/swdi/stormevents/csvfiles/"

  # Extend the data back to 2016 to get a longer view of trends
  years = 2016:2024

  RAW_DIR = file.path("data", "raw")
  dir.create(RAW_DIR, recursive = TRUE, showWarnings = FALSE)

  # Read the html code for the NOAA website so I can get the files
  storm_index = read_html(base_url)

  # Pull only the links, keep only those with the yearly details pattern
  storm_files = storm_index |>
    # Extract all HTML link elements <a> from the page
    html_elements("a") |>
    # Get the href attribute from each link, which contains the file name
    html_attr("href") |>
    # Turn the vector of file names into a tibble with one column called file
    tibble(file = _) |>
    # Keep files that match the NOAA Storm Events data format using Regex
    # This pattern looks for names like:
    # StormEvents_details-ftp_v1.0_d2024_c20260323.csv.gz
    filter(str_detect(file,
                    "^StormEvents_details-ftp_v1\\.0_d\\d{4}_c\\d{8}\\.csv\\.gz$")) |>
    # Extract the 4-digit year from the file name and store it as an integer
    mutate(year = as.integer(str_extract(file, "(?<=d)\\d{4}(?=_c)"))) |>
    # Keep only the years selected above
    filter(year %in% years) |>
    # Keep only the years 2016-2024
    arrange(year)

  # Function that downloads one NOAA file at a time
  download_one = function(file_name, dest_dir = RAW_DIR, pause = 1.5) {
    # Create a succinct local file path where this file will be saved
    dest = file.path(dest_dir, file_name)

    # Only download the file if it is not already saved locally
    if (!file.exists(dest)) {
      tryCatch(
        download.file(
          # Build the full web address for this specific file
          url = paste0(base_url, file_name),
          # Save the downloaded file to the destination path
          destfile = dest,
          # Binary writing mode, for compressed files
          mode = "wb",
          # Don't output anything
          quiet = TRUE
        ),
        error = function(e) {
          message("  Download failed: ", conditionMessage(e))
          # Remove any partial file that was created
          if (file.exists(dest)) file.remove(dest)
        }
      )
      # Pause requests to avoid rate limit issues
      Sys.sleep(pause)
    }
    # Return the file path
    dest
  }

  # Now use the funtion to every NOAA file
  storm_paths = map_chr(storm_files$file, download_one)

  # Keep only paths that actually exist in case any downloads failed
  storm_paths_valid = storm_paths[file.exists(storm_paths)]

  # Read all downloaded files into R and combine them into one large data frame,
  # this will not work everywhere, NOAA is fortunately consistent with data structure
  # I read every column as character to avoid parsing problems across years
  raw_combined = map_dfr(
    storm_paths_valid,
    ~ read_csv(.x, col_types = cols(.default = col_character()),
               show_col_types = FALSE)
  )

  # Save the combined raw dataset as a CSV file so I can inspect it later
  # without needing to re-download
  write_csv(raw_combined, combined_cache)
}

# Create a vector of the severe weather event types that I want to keep
event_types_keep = c("Tornado", "Hail", "Thunderstorm Wind", "Flash Flood")

# Filter the data to only include only the event types I want to examine
# Also covert fields into their correct types
severe_events = raw_combined |>
  select(
    BEGIN_DATE_TIME, EVENT_TYPE, STATE,
    BEGIN_LAT, BEGIN_LON, INJURIES_DIRECT, DEATHS_DIRECT
  ) |>
  # Use %in% to keep only those event types
  filter(EVENT_TYPE %in% event_types_keep) |>
  mutate(

    # Covert dates to a correct date-time format
    begin_dt = ymd_hms(BEGIN_DATE_TIME),

    # Extract the year
    year = year(begin_dt),

    # Extract the month for the variability analysis
    month = month(begin_dt),

    # Convert latitude and longitude from character to numeric so they can
    # be filtered
    BEGIN_LAT = as.numeric(BEGIN_LAT),
    BEGIN_LON = as.numeric(BEGIN_LON),

    # Convert injuries and deaths from character to numeric
    INJURIES_DIRECT = as.numeric(INJURIES_DIRECT),
    DEATHS_DIRECT = as.numeric(DEATHS_DIRECT)
  ) |>

  # Added another layer of filters to make sure required fields are correct
  filter(

    # Keep only rows where the date-time was successfully parsed
    !is.na(begin_dt),

    # Keep only rows with valid longitude and latitude values
    !is.na(BEGIN_LON),
    !is.na(BEGIN_LAT),

    # Restrict the data to the approximate bounds of the continuous
    # United States
    between(BEGIN_LON, -125, -66),
    between(BEGIN_LAT, 24, 50)
  ) |>

  # Keep all the cleaned stuff
  select(
    EVENT_TYPE, STATE, begin_dt, year, month,
    BEGIN_LAT, BEGIN_LON,
    INJURIES_DIRECT, DEATHS_DIRECT
  )
```

# About

## Row

```{r}
#| content: valuebox
#| title: "Total Events (2016-2024)"
list(
  icon = "cloud-lightning-rain",
  color = "primary",
  value = format(nrow(severe_events), big.mark = ",")
)
```

```{r}
#| content: valuebox
#| title: "Years Covered"
list(
  icon = "calendar-range",
  color = "secondary",
  value = "2016 - 2024"
)
```

```{r}
#| content: valuebox
#| title: "States Included"
list(
  icon = "geo-alt",
  color = "info",
  value = n_distinct(severe_events$STATE)
)
```

## Row

### About This Dashboard

My goal is to investigate whether severe weather in the US appears to be becoming more volatile over time. By volatile, I mean not just more frequent on average, but more variable from year to year and harder to predict. This dashboard uses the NOAA Storm Events database to explore four severe convective weather event types from 2016 to 2024. This work is part of a larger research project where I plan to use machine learning and operations research for forecasting.

The NOAA Storm Events database contains records of significant weather events across the United States going back to 1950. Each row represents a single event, with fields for event type, start date and time, state, location, injuries, deaths, and property damage. I focus on four event types in the contiguous United States: **Tornado**, **Hail**, **Thunderstorm Wind**, and **Flash Flood**.

Use the **Event Trends** page to explore how annual event counts and state-level distributions have changed over time. Use the **Variability** page to look at within-year and year-to-year variability. Both pages have filters so you can focus on specific event types and year ranges. The **Summary** page has key findings and links to additional resources.

# Event Trends

## {.sidebar}

**Filters**

```{r}
# Let the user select which event types to display
checkboxGroupInput(
  "trend_types",
  "Event Types:",
  choices = event_types_keep,
  selected = event_types_keep
)

# Let the user narrow the year range
sliderInput(
  "trend_years",
  "Year Range:",
  min = 2016,
  max = 2024,
  value = c(2016, 2024),
  step = 1,
  sep = ""
)
```

## Row

### Annual Event Counts

```{r}
plotOutput("trendCountsPlot")
```

## Row

### Top 10 States by Number of Events

```{r}
plotOutput("trendStatesPlot")
```

# Variability

## {.sidebar}

**Filters**

```{r}
# Let the user select which event types to display
checkboxGroupInput(
  "var_types",
  "Event Types:",
  choices = event_types_keep,
  selected = event_types_keep
)

# Let the user narrow the year range
sliderInput(
  "var_years",
  "Year Range:",
  min = 2016,
  max = 2024,
  value = c(2016, 2024),
  step = 1,
  sep = ""
)
```

## Row

### Monthly Event Count Distribution by Year

```{r}
plotOutput("varMonthlyPlot")
```

## Row

### Annual Variability Summary

```{r}
tableOutput("varTable")
```

### Annual Trends with Linear Fit

```{r}
plotOutput("varTrendPlot")
```

# Summary

## Row

### Key Findings

Looking at NOAA Storm Events data from 2016 to 2024, a few patterns stand out. Tornado counts have the highest coefficient of variation across all four event types, meaning they are the most erratic relative to their average level even though they are the least common event type in absolute terms. Thunderstorm wind events are the most common by a wide margin but are also the most stable from year to year. Texas leads all states in total severe weather reports across all event types, though some of this reflects its size.

The linear trend lines on the Variability page show mostly flat or slightly changing slopes, but the confidence bands are wide. With only nine years of data, it is hard to make strong statistical claims from the visual trends alone. A more formal test using a Bayesian Poisson GLM is the next step in this analysis.

This dashboard is part of a larger research project investigating whether physics-constrained neural networks can improve severe weather forecasting under non-stationary climate conditions.

### Additional Resources

- [NOAA Storm Events Database](https://www.ncei.noaa.gov/pub/data/swdi/stormevents/csvfiles/) — Raw data source used in this project
- [NOAA Storm Data Documentation](https://www.nws.noaa.gov/directives/sym/pd01016005curr.pdf) — Field definitions and methodology for the Storm Events dataset
- [SPC Severe Weather Climatology](https://www.spc.noaa.gov/wcm/) — NOAA Storm Prediction Center historical data
- [Gensini and Brooks (2018)](https://www.nature.com/articles/s41612-018-0048-2) — Spatial trends in US tornado frequency
- [Brooks (2013)](https://www.sciencedirect.com/science/article/pii/S0169809512003153) — Severe thunderstorms and climate change

```{r}
#| context: server

# Server logic for Event Trends page

# Create a line plot that shows how many of each event occurred a year
output$trendCountsPlot = renderPlot({
  severe_events |>
    filter(
      EVENT_TYPE %in% input$trend_types,
      between(year, input$trend_years[1], input$trend_years[2])
    ) |>
    count(year, EVENT_TYPE) |>
    ggplot(aes(x = year, y = n, color = EVENT_TYPE)) +
    geom_line(linewidth = 1) +
    geom_point(size = 2) +
    scale_x_continuous(breaks = input$trend_years[1]:input$trend_years[2]) +
    labs(
      title = "Severe Weather Events by Year",
      x = "Year",
      y = "Number of Events",
      color = "Event Type"
    ) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
})

# Create a horizontal bar chart that looks at how many sever events in
# general occur for each state in the US
output$trendStatesPlot = renderPlot({
  severe_events |>
    filter(
      EVENT_TYPE %in% input$trend_types,
      between(year, input$trend_years[1], input$trend_years[2])
    ) |>
    # Just in case there are some obs with no listed state
    filter(!is.na(STATE), STATE != "") |>
    # Sorted so longest bar is on top
    count(STATE, sort = TRUE) |>
    # Only look at the top ten states
    slice_head(n = 10) |>
    ggplot(aes(x = reorder(STATE, n), y = n)) +
    geom_col(fill = "steelblue") +
    coord_flip() +
    labs(
      title = "Top 10 States by Number of Recorded Events",
      x = "State",
      y = "Number of Events"
    ) +
    theme_minimal()
})

# Server logic for Variability page

# Create a boxplot showing the distribution of monthly counts for each year
# Wider boxes indicate more spread across months in that year
output$varMonthlyPlot = renderPlot({
  severe_events |>
    filter(
      EVENT_TYPE %in% input$var_types,
      between(year, input$var_years[1], input$var_years[2])
    ) |>
    # Count events by year, month, and event type
    count(year, month, EVENT_TYPE) |>
    ggplot(aes(x = factor(year), y = n)) +
    geom_boxplot(outlier.shape = 1, fill = "lightsteelblue", color = "steelblue4") +
    facet_wrap(~EVENT_TYPE, scales = "free_y", ncol = 2) +
    labs(
      title = "Monthly Event Count Distribution by Year",
      subtitle = "Wider boxes indicate greater within-year variability",
      x = "Year",
      y = "Events per Month"
    ) +
    theme_minimal() +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1),
      strip.text = element_text(face = "bold")
    )
})

# Summarize annual event counts and compute the coefficient of variation
output$varTable = renderTable({
  severe_events |>
    filter(
      EVENT_TYPE %in% input$var_types,
      between(year, input$var_years[1], input$var_years[2])
    ) |>
    count(year, EVENT_TYPE) |>
    group_by(EVENT_TYPE) |>
    summarise(
      mean_annual = round(mean(n), 0),
      sd_annual = round(sd(n), 1),
      cv = round(sd(n) / mean(n), 3),
      .groups = "drop"
    ) |>
    arrange(desc(cv)) |>
    rename(
      `Event Type` = EVENT_TYPE,
      `Mean Annual Count` = mean_annual,
      `Std Dev` = sd_annual,
      `CV` = cv
    )
})

# Create a trend plot with linear fits and confidence bands for each event type
output$varTrendPlot = renderPlot({
  severe_events |>
    filter(
      EVENT_TYPE %in% input$var_types,
      between(year, input$var_years[1], input$var_years[2])
    ) |>
    count(year, EVENT_TYPE) |>
    ggplot(aes(x = year, y = n, color = EVENT_TYPE, fill = EVENT_TYPE)) +
    geom_line(linewidth = 1) +
    geom_point(size = 2) +
    geom_smooth(method = "lm", se = TRUE, alpha = 0.12, linewidth = 0.6) +
    scale_x_continuous(breaks = input$var_years[1]:input$var_years[2]) +
    labs(
      title = "Annual Event Counts with Linear Trend",
      subtitle = "Shaded bands are 95% confidence intervals on the linear fit",
      x = "Year",
      y = "Annual Event Count",
      color = "Event Type",
      fill = "Event Type"
    ) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
})
```
